---
title: "Global Catch data for 2014"
output: 
   html_document:
      number_section: yes
      toc: yes
      theme: united
author: "Thomas Berger"
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Tis project produces a map of Global fisheries, showing the official catch data (FAO Global Capture Production) for the most recent year available, which was 2014 when this report was created.

Some species are reported by weight (tons), while others are reported by number (no): for the color coding only the weight is considered; while the hover popup shows also the detail data (name of country, catch in tons).

This is the markdown document contains all steps to create the output for this project.

The data was loaded from the FAO query panel:
http:www.fao.org/fishery/statistics/global-capture-production/query/en, and saved into the file 'figis_guestnull.csv'. To arrive at the detail, select country [Alphabetical], on the Index select [All] to have all countries, then click [select all] to include them all into the query. Under time, select [2014] as year. Under Species, select [Ybk main groups], select [Fish, crustaceans and molluscs, etc.] to have data for human consumption. Under Display, select [Country] and [Quantity] to have the result by country. Save the data using the [Export] button.

FAO.FishStat.Global Capture Production.Data: 2014.<br>
FAO - Fisheries and Aquaculture Information and Statistics Branch - downloaded 09/01/2017

The github repository is here:
https://github.com/bergertom/GlobalCapture

The markdown result can be accessed here:
https://bergertom.github.io/GlobalCapture/ReadMe.html

And the leaflet is published here:
https://bergertom.github.io/GlobalCapture/

## Loading and cleaning the data

```{r }
   # read data downloaded fom FAO web site
   faodata <- read.csv("figis_guestnull.csv", na.strings = c("NA", "", NULL))
   # load the country mapping to ISO3
   country <- read.csv("country.csv", na.strings = c("NA", "", NULL))
   
   # show the dimensions
   str(faodata)
   str(country)
```

```{r }
   # change column names
   colnames(faodata)<- c("Country","Unit","Value.2014","Symbol.2014")
   colnames(country)<- c("ISO3","Country")
   
   # get rid of statistical symbol
   faodata$Symbol.2014 <- NULL
```


```{r }
   # merge with ISO3 codes required for interactive map
   faodata <- merge(faodata, country, by = "Country")
   # remove numbers
   faodata <- subset(faodata, Unit == 't')
   # remove unit
   faodata$Unit <- NULL
```

## Creating the Leaflet

The first step is to add the popup text to the data
```{r }
   # create the data for the popup
   faodata$popup <- paste0("<strong>Country: </strong>", 
                        faodata$Country, " (", faodata$ISO3, ")",
                        "<br><strong>", 
                        "Catch (tons)", 
                        ": </strong>", 
                        round(faodata$Value.2014),0)
```

Download the map data
```{r }
   library(rgdal)

   # We download the data for the map from naturalearthdata.com
   url <- "http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/50m/cultural/ne_50m_admin_0_countries.zip"
   folder <- getwd() #set a folder where to download and extract the data
   file <- basename(url) 
   if (!file.exists("ne_50m_admin_0_countries.shp")) {
       # download and unzip the country maps only once
      download.file(url, file)
      unzip(file, exdir = folder)
   }

   # then read it with rgdal library
   world <- readOGR(dsn = folder, 
                     layer = "ne_50m_admin_0_countries",
                     encoding = "UTF-8", #you may need to use a different encoding
                     verbose = FALSE)

   # merge the polygons with the FAO data
   world2 <- merge(world, faodata,
               by.x = "adm0_a3",
               by.y = "ISO3",
               sort = FALSE)
   # remove all contries where there is no data (ISO3 match can be improved)
   world2 <- world2[!is.na(world2$Value.2014),]
```

Create the interactive map:
```{r }
   library(leaflet)
   #Colour palette. Check ?colorQuantile and ?RColorBrewer for more options
   pal <- colorQuantile("YlOrRd", NULL, n = 20)
   #Tiles coming from stamen.com
   tiles <- "http://{s}.tile.stamen.com/toner-lite/{z}/{x}/{y}.png"
   # legal copyrights
   attribution <- paste0('by Thomas Berger generated on ',format(Sys.Date(), "%Y-%m-%d"),
        ', Maps by <a href="http://www.naturalearthdata.com/">Natural Earth</a>',
        ', 2014 Fishery capture data by <a href="http:www.fao.org/fishery/statistics/">FAO</a>.')
   
   #And finally the map:
   ll <- leaflet(data = world2) %>%
      addTiles(urlTemplate = tiles, attribution = attribution) %>%
      setView(0, 0, zoom = 2) %>%
      addPolygons(fillColor = ~pal(world2$Value.2014), 
              fillOpacity = 0.8, 
              color = "#000000", 
              weight = 1, 
              popup = world2$popup)
   ll
   
   # save to stand-alone html file
   library(htmlwidgets)
   saveWidget(ll, file="index.html")
```


## References
* Country maps http://www.naturalearthdata.com/features/
* FAO query panel http:www.fao.org/fishery/statistics/global-capture-production/query/en

